---
title: "Threats Assessment Methods for NBW Trends"
author: "Author"
date: "2025"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup , include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      fig_caption= TRUE,
                      fig.align = "left",
                      fig.pos = 'h',
                      dev = "png", 
                      cache = F
                )
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r libraries, include=FALSE}
#libraries
pacman::p_load(knitr, dplyr, sf, viridis, ggplot2, 
stars,
stringr,
here,
fasterize,
ggtext,
ggspatial,
pals,
tidyr,
fuzzyjoin,
kableExtra)

#read in functions
source(here::here("scripts/functionsCHI.r"))

options(scipen = 999) 

```

------------------------------------------------------------------------

## Data Processing {.tabset}

### 1. Import basemap data:

-   This includes Coastlines, bathymetry, NBW habitat area, and
    conservation zones
-   Code imports data and transforms to UTM Zone 20 where necessary
-   Extent of study area is based on NBW important habitat from Feyrer et al. 2024
-   Bathymetric data from Atlas of Canada available at:
    (<https://ftp.maps.canada.ca/pub/nrcan_rncan/vector/framework_cadre/North_America_Atlas10M/bathymetry/>)
-   Conservation zones compiled from layers available as shapefiles.
    -   Ocean's Act MPAs from
        <https://open.canada.ca/data/en/dataset/a1e18963-25dd-4219-a33f-1a38c4971250>\
    -   Northern bottlenose whale Critical Habitat designations from
        Species at Risk
        <https://open.canada.ca/data/en/dataset/db177a8c-5d7d-49eb-8290-31e6a45d786c>

```{r import_basemap, echo = FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE}
source(here::here("scripts/baseMap2.r"), local = knitr::knit_global())
#remove any previous results before knitting and rerunning all scripts
unlink(here::here("output/GRIDS/*"))
#map base map figure 1

```

------------------------------------------------------------------------

### 2. Map threats based on relative intensity of human activities

-   The spatial distribution of serious threats across the study area
    was interpreted from available information on the intensity of
    associated human activities: Entanglement, Depredation (Fishing
    effort), Military Sonar (Military Exercise Zones), Plastics (Density
    of large debris and micro plastics), Large Oil spills (Oil and Gas
    Development) and Vessel strikes (Vessel Traffic).
-   Whaling has not occurred in the region across the study period.
-   Data is divided into two periods, historical (1988-2004) and
    contemporary (2005-2023). Information on the intensity of human
    activities was standardized to deciles (e.g., 10% bins).

#### A. Fishing Effort (Entanglement, Depredation)

-   Fishing effort was assessed using a combination of publicly
    available intensity data (landings or time spent in each cell) for the region
    for three primary gear classes (i) Groundfish Fixed gear, e.g.,
    bottom longline, (ii) Pelagic longline, and (ii) Groundfish Mobile
    gear, e.g. trawl fisheries.
-   Differences in the methods for assessing effort between fisheries
    and time periods result in some differences in the resolution of
    available information. All effort data was standardized into
    deciles.

```{r Pre/Post Fishing effort, include = TRUE, fig.fullwidth = TRUE, message=FALSE, warning=FALSE}
source(here::here("scripts/CompareFish.r"), local = knitr::knit_global())

```

#### B. Military exercise zones (Military Sonar)

-   Military exercises– The Department of National Defence has
    designated Firing Practice and Exercise Areas off the coasts of
    Canada. Activities in these areas may include bombing practice from
    aircraft, air-to-air, air-to-sea or ground firing, and anti-aircraft
    firing, etc. In Atlantic Canada, this includes sea area employments
    for sub-surface operations and firing exercises (FIREX) (Department
    of National Defense Government of Canada, 2021). Polygon shapefiles
    of the designated areas were downloaded from the Canadian Government
    open data portal
    (<https://open.canada.ca/data/en/dataset/73111c78-298b-4be9-97f1-7aaa73cab477>).

-   With no effort data on military exercises in these areas across the
entire period, we interpreted them to represent threat potential for military
mid-frequency active sonar (MMFAS). Although noise impacts can extend outside
these areas, the most acute threat would occur inside the exercise areas. There
is little information on the extent of military activity in the early period,
however the Canadian Coast Guard Navigational Warnings (NAVWARNs), previously
known as Notices to Shipping (NOTSHIPS), indicate there were 44 notices of
military exercises in areas overlapping with beaked whale habitat, on 54 days in
2019 (the first year notices were made available online), lasting between 1-5
days each (Government of Canada, 2021). However, this included 8
(non-consecutive) days of the biannual international CUTLASS FURY exercises.
This biannual event began in 2016 and has been described as the “largest
international military exercise in Canadian waters” (The Canadian Press, 2016).
Excluding CUTLASS FURY 2019, there were 46 days per year with notifications to
mariners (12.6% year). We removed the MPA from this layer in the contemporary
period, as the MPA management plan (Fisheries and Oceans Canada, 2017) indicates
that although the Gully MPA cannot legally exclude military activities, in
practise the government agrees that such exercises do not occur inside the MPA.

```{r DND, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE, fig.fullwidth = TRUE}

source(here::here("scripts/militaryexMap.r"), local = knitr::knit_global())

#Early Period (1988-2004)------# # Contemporary Period (2005-2019)-------
# Plot map
m8+m9 + plot_layout(guides = "collect")& theme(legend.position = 'bottom')

```

#### C. Oil and Gas Development (Large Oil spills)

-   Publicly available spatial and temporal data was collated from the
    Canada-Nova Scotia Offshore Petroleum Board (C-NSOPB) websites
    (CNSOPB, 2021). Shapefiles for existing pipelines, wells, platforms,
    exploratory licenses, significant discovery licenses, production
    licenses and sector announcements provided as part of the call for
    bids processes were downloaded, cleaned, and checked for
    duplication.

-   Activities occurring before and after 2004 were identified based on
    metadata in the shapefile or provided elsewhere on the website.

    -   Oil & Gas Operations includes wells, pipelines, and platforms
    -   Some infrastructure has been decommissioned since it was built,
        however the ongoing risk associated with pollution is assumed to
        continue over the entire study period.

```{r O&G operations, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE}

 source(here::here("scripts/oilngasMap.r"), local = knitr::knit_global())

#Early Period (1988-2004)------# # Contemporary Period (2005-2019)-------
# Plot maps
m10+ m11 + plot_layout(guides = "collect")& theme(legend.position = 'bottom')
```

#### D. Commercial Vessel Traffic (Vessel Strike)

-   Intensity raster of global shipping traffic for early period based
    on Scotian Shelf Human Use Atlas which provides a description of annual
    shipping patterns and intensity during 2002. Available data includes
    only inbound traffic; however, the density pattern of departing
    vessels is much the same as the inbound traffic pattern.
-   Contemporary shipping data is based on 2019 daily intensity shipping (OpenGov.ca), reflecting pre-covid shipping patterns. Increased shipping around Sable Island reflects the decommissioning of Oil and Gas infrastructure between 2017-2019. https://www.cnsopb.ns.ca/offshore-activity/current-activity/sable-offshore
-   There is a voluntary vessel avoidance area around the Gully MPA,
    however there is no regulatory exclusion of this threat.

```{r Shipping Effort, echo= FALSE, warning=FALSE, message=FALSE, fig.fullwidth=T, message=T, warning=T}
source(here::here("scripts/shipMap.r"), local = knitr::knit_global())

m14 + m15 + plot_layout(guides = "collect")& theme(legend.position = 'bottom')

```

### 3. Assessing difference in intensity of threats

-   Scale intensity data so values range between 0-100 for activities with
    effort information: Fishing effort, Shipping,
-   Presence /absence footprint data is either 0 or 1: Oil & Gas Explorations, Military Exercise areas
-   The relative intensity of activities in each period are binned into deciles to
    visualize areas of increased and decreased intensity of potential threats

```{r Calc threat dif, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE }
source(here::here("scripts/calcThreatDif.r"), local = knitr::knit_global(), )

# plot by threat
threats2_combined

#plot dif btwn post-pre
diffThreats_plot

```

