---
title: "Cumulative Human Impact (CHI) scripts Methods"
author: "Feyrer et al. 2024"
date: "July 2024"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup , include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      fig_caption= TRUE,
                      fig.align = "left",
                      fig.pos = 'h',
                      dev = "png", 
                      cache = F
                )
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r libraries, include=FALSE}
#libraries
pacman::p_load(knitr, dplyr, sf, viridis, ggplot2, 
stars,
stringr,
here,
fasterize,
ggtext,
ggspatial,
pals,
tidyr,
fuzzyjoin,
kableExtra)

#read in functions
source(here::here("scripts/functionsCHI.r"))

options(scipen = 999) 

```

------------------------------------------------------------------------

## Data Processing {.tabset}

### 1. Import basemap data:

-   This includes Coastlines, bathymetry, NBW habitat area, and
    conservation zones
-   Code imports data and transforms to UTM Zone 20 where necessary
-   Extent of study area is based on area of influence around NBW
    habitat being a 50km buffer of the 1000m bathy line
-   Bathymetric data from Atlas of Canada available at:
    (<https://ftp.maps.canada.ca/pub/nrcan_rncan/vector/framework_cadre/North_America_Atlas10M/bathymetry/>)
-   Conservation zones compiled from layers available as shapefiles.
    -   Other Effective Conservation Measures (OECM) from
        <https://open.canada.ca/data/en/dataset/44769543-7a23-4991-a53f-c2cf7c7a946f>\
    -   Ocean's Act MPAs from
        <https://open.canada.ca/data/en/dataset/a1e18963-25dd-4219-a33f-1a38c4971250>\
    -   Georges Bank Oil and Gas Exclusion zone from CNSOPB
        <https://callforbids.cnsopb.ns.ca/2016/01/data-environment/gis-information>\
    -   Northern bottlenose whale Critical Habitat designations from
        Species at Risk
        <https://open.canada.ca/data/en/dataset/db177a8c-5d7d-49eb-8290-31e6a45d786c>

```{r import_basemap, include=TRUE, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 5, fig.fullwidth = TRUE}
source(here::here("scripts/baseMap.r"), local = knitr::knit_global())

#remove any previous results before knitting and rerunning all scripts
unlink(here::here("Results/GRIDS/scaled/*"))
#map base map

include_graphics(gg_Fig2path)

```

------------------------------------------------------------------------

### 2. Import stressor data

-   Data is divided into two periods, early (1988-2004) and contemporary
    (2005-2019)
-   Stressors with regionally specific information are Longline Fishing
    effort, Military Exercises, Oil and Gas Exploration and Development
    Activities\
-   Remaining stressors (Other Fishing gears, Shipping and SST
    anomalies) are sourced from Halpern et al. 2015

#### A. Fishing Effort

-   Long line fishery effort was not available as part of Halpern et
    al.'s global data set for the study area. Code below imports data
    for the fisheries created by:
-   S. Butler, D. Ibarra and S. Coffen-Smout, 2019. Maritimes Region
    Longline and Trap Fisheries Footprint Mapping for Marine Spatial
    Planning and Risk Assessment. Can. Tech. Rep. Fish. Aquat. Sci.
    3293: v + 30 p. Data available at:
    <https://open.canada.ca/data/en/dataset/3d2e1a84-20f5-4a61-90e4-94760c80ebb9>

##### i. Bottom Longline effort

-   Uses bottom long line fishing effort data created by Butler et al.
    2019, where fishing effort is sum of effort in hrs/ km2 over the
    entire period 2002-2017.
-   Early Period uses this dataset as is, given overall spatial effort
    has not changed. Contemporary period uses this dataset with Zone 1
    of Gully removed to account for fishing closures that occured there
    post 2004.

```{r Butler Bottom Longline, include = TRUE, fig.fullwidth = TRUE}
source(here::here("scripts/fishMap.r"), local = knitr::knit_global())

#Bottom Longline
# Plot map

m2+m3+ plot_layout(guides = "collect")& theme(legend.position = 'bottom')

```

##### ii. Pelagic longline effort

-   Pelagic longline fishing effort data is provided in mins/ km2 from
    2003-2018 in log scale.
-   Dataset focuses on vessels reporting to DFO Maritimes region and
    only covers the western region off Newfoundland. Additional effort
    occurs in Newfoundland waters but was not included in this dataset.
-   Early Period uses this dataset as is, given overall spatial effort
    has not changed. Contemporary period uses this dataset with Zone 1
    of Gully removed to account for fishing closures that occured post
    2004.

```{r Butler Pelagic Longline, include = TRUE, fig.fullwidth = TRUE}

#Pelagic Longline
# Plot map

m4+m5+ plot_layout(guides = "collect")& theme(legend.position = 'bottom')

```

##### iii. Other Fishing Effort

-   Halpern et al. (2015) used landings normalized globally by
    productivity for different gear types
-   Methods for estimating fishing effort is based annual wild caught
    industrial fisheries catch (tonnes) by taxa and gear type at 0.5°
    resolution as described in:
-   Halpern, B. S. et al. Patterns and emerging trends in global ocean
    health. PlosOne <DOI:10.1371/journal.pone.0117863> (2015).

Layers are separated based on gear type and in units of tons of fish per
tons of carbon in each 1 km2. Data downloaded from:\
Benjamin Halpern, Melanie Frazier, John Potapenko, Kenneth Casey, Kellee
Koenig, et al. 2015. Cumulative human impacts: raw stressor data (2008
and 2013). Knowledge Network for Biocomplexity. <doi:10.5063/F1S180FS>.

Gear type layers included here: i. demersal destructive (e.g., bottom
trawl), ii. demersal non-destructive high bycatch (e.g., pots, traps),
iii. demersal non-destructive low bycatch (e.g., hook and line), iv.
pelagic low bycatch (e.g., hook and line) - Post 2004, consistent with
other fishing effort layers uses this dataset with Zone 1 of Gully
removed to account for fishing closures that occured in 2005

```{r Halpern et al. Fishing, include = TRUE, fig.fullwidth = TRUE}

#Early Period (1988-2004)------# # Contemporary Period (2005-2019)-------
# Plot map
m12 +m13 + plot_layout(guides = "collect")& theme(legend.position = 'bottom')

```

#### B. Military activities (MFAS)

-   Military exercises– The Department of National Defence has
    designated Firing Practice and Exercise Areas off the coasts of
    Canada. Activities in these areas may include bombing practice from
    aircraft, air-to-air, air-to-sea or ground firing, and anti-aircraft
    firing, etc. In Atlantic Canada, this includes sea area employments
    for sub-surface operations and firing exercises (FIREX) (Department
    of National Defense Government of Canada, 2021). Polygon shapefiles
    of the designated areas were downloaded from the Canadian Government
    open data portal
    (<https://open.canada.ca/data/en/dataset/73111c78-298b-4be9-97f1-7aaa73cab477>).

-   With no effort data on military exercises in these areas across the
    entire period, we interpreted them to represent threat potential for
    military mid-frequency active sonar (MMFAS). Although noise impacts
    can extend outside these areas, the most acute threat would occur
    inside the exercise areas. According to the Canadian Coast Guard
    Navigational Warnings (NAVWARNs), previously known as Notices to
    Shipping (NOTSHIPS), there were 44 notices of military exercises in
    areas overlapping with beaked whale habitat, on 54 days in 2019 (the
    first year notices were made available online), lasting between 1-5
    days each (Government of Canada, 2021). However, this included 8
    (non-consecutive) days of the biannual international CUTLASS FURY
    exercises. This biannual event began in 2016 and has been described
    as the “largest international military exercise in Canadian waters”
    (The Canadian Press, 2016). Excluding CUTLASS FURY 2019, we used the
    proportion of days per year with notifications to mariners (46 days
    = 12.6%) as a baseline for effort across the area in any given year.
    Although there is little information on the extent of military
    activity in the early period, there were two large international
    exercises as part of the public record in 2016 and 2019 (i.e.,
    CUTLASS FURY). To account for the impact of the international
    exercises, we have increased the % effort to 15% per year in the
    period after 2004. In addition, we removed the MPA from this layer
    in the contemporary period, as the MPA management plan (Fisheries
    and Oceans Canada, 2017) indicates that although the Gully MPA
    cannot legally exclude military activities, in practise the
    government agrees that such exercises do not occur inside the MPA.

```{r DND, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE, fig.fullwidth = TRUE}

source(here::here("scripts/militaryexMap.r"), local = knitr::knit_global())

#Early Period (1988-2004)------# # Contemporary Period (2005-2019)-------
# Plot map
m6+m7 + plot_layout(guides = "collect")& theme(legend.position = 'bottom')

```

#### C. Oil & Gas Exploration and Operations

-   Publicly available spatial and temporal data was collated from the
    Canada-Nova Scotia Offshore Petroleum Board (C-NSOPB) and the
    Canada-Newfoundland and Labrador Offshore Petroleum Board (C-NLOPB)
    websites (CNLOPB, 2021; CNSOPB, 2021). Shapefiles for existing
    pipelines, wells, platforms, seismic surveys (2D/3D), exploratory
    licenses, significant discovery licenses, production licenses and
    sector announcements provided as part of the call for bids processes
    were downloaded, cleaned, and checked for duplication.
-   Activities occurring before and after 2004 were identified based on
    metadata in the shapefile or provided elsewhere on the website.
-   The extent of oil and gas development activities in the portion of
    the study area regulated by the C-NLOPB (i.e., off Newfoundland) may
    be incomplete but this is unclear due to current data embargoes.
    However, based on our review of public reports on activities in the
    area, likely most seismic surveys were included in our scripts.

##### i. Exploratory Activities

-   Oil & Gas Exploration includes seismic surveys, exploratory licenses
    and licensed activities such as shipping and test wells
-   Data includes seismic surveys and exploratory licenses
-   Individual seismic survey data only available for period up to 2004.
    Shapefiles for survey data was deduplicated, merged and shapefiles
    clipped to SDM area. Seismic surveys are represented as either 3D
    (polygons) or 2D (lines)
-   Exploratory Licenses have a 9 year term and post 2004 are only data
    available on exploratory activities

```{r O&G explore, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE}

 source(here::here("scripts/oilngasMap.r"), local = knitr::knit_global())

#Early Period (1988-2004)------# # Contemporary Period (2005-2019)-------
# Plot maps
m8+ m9 + plot_layout(guides = "collect")& theme(legend.position = 'bottom')

```

##### ii. Operational activities

-   Oil & Gas Operations includes wells, pipelines, and platforms which
    are treated as presence absence data as per Halpern et al. (2008)
-   Some of this infrastructure has been decommissioned since it was
    built, however the ongoing risk associated with pollution is assumed
    to continue over the entire study period.

```{r Operations, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE}
#Early Period (1988-2004)------# # Contemporary Period (2005-2019)-------

# Plot maps
m10+ m11 + plot_layout(guides = "collect")
```

#### D. Shipping

-   From Halpern et al. 2015. Intensity raster of global shipping
    traffic is the same for both periods
-   As described in Halpern et al. 2015, this layer combines several
    years of shipping data but are mostly a static description of
    shipping patterns and intensity during 2011.
-   Shipping is not excluded from Gully MPA

```{r Shipping Effort, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE}
source(here::here("scripts/shipMap.r"), local = knitr::knit_global())

m14

```

#### E. Climate Change - Positive SST anomalies

-   From Halpern et al. 2015. Number of positive SST anomalies in the
    period relative to SST in 1985-1990.
-   From Halpern et al. 2008 "Changes in sea surface temperature as a
    result of climate change will have varying effects on species and
    ecosystems... Potential for ecological change in response to changes
    in sea surface temperature can be measured as the frequency of
    temperature anomalies, where the temperature exceeds a threshold
    value like the long-term mean, or by a measure of the magnitude (°C)
    of the anomalies themselves (S19, 20)."

```{r Climate change, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE}
source(here::here("scripts/sstMap.r"), local = knitr::knit_global())

#Early Period (1988-2004)------# # Contemporary Period (2005-2019)-------
# Plot maps
m15+m16 + plot_layout(guides = "collect")& theme(legend.position = 'bottom')
```

### 3. Assessing cumulative impacts from individual threats

-   Scale stressor intensity data so values range between 0-1 for layers
    with effort information: Oil & Gas Explorations, Fishing effort,
    Shipping, SST anomalies
-   Read in Pre/Post layers, Log10[x+1] transform and then normalize
    effort grids by max value across both periods
-   Presence /absence of stressor footprint data values should be either
    0 or 1
-   Stressor layers (effort or footprint) for each of the two periods
    are multiplied by the NBW vulnerability score
-   NBW Vulnerability score assigned based on sensitivity, temporal
    frequency and extent of stressor (See Impact Score Rationale below,
    Table 2 in Feyrer et al. 2022)
-   Finally the values of stressors in each period are summed to assess
    Cumulative impact

```{r Calc CHI, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE }
source(here::here("scripts/rescaleStressor.r"), local = knitr::knit_global())
source(here::here("scripts/calcCHI.r"), local = knitr::knit_global(), )

# plot by threat
include_graphics(fig3threatspath)
ImpactScore1 = ImpactScore[c(1,3,2, 4:6), 2:6]
names(ImpactScore1) <- gsub('[.].+', '', names(ImpactScore1))

#Table 1
 #format table
  kbl(ImpactScore1, booktabs = TRUE, caption = "Impact Score Rationale Table" ) %>%kable_styling(latex_options="scale_down")%>% row_spec (0, bold = T)%>%landscape( margin = NULL)

  
  #plot CHI PRE/ POST
include_graphics(fig4CHIpath)
```

### 4. Calculating difference in CHI between periods

-   The difference in Cumulative impact between periods highlights
    measures the cell-by-cell difference between CHI
-   Can be used to understand areas and intensity of differences in
    threats to NBW over time
-   Absolute diference is Contemporary period - Early Period

##### RMS Error

-   Square Root(sum(e\^2)/n)
-   About 2/3 of all the cells will differ by less than the rmse.
-   About 95% of all the cells will differ by less than twice the rmse.

```{r Calc CHI difference, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE}

# Plot Absolute Difference between periods
# Plot RMSE Difference & Plot RMSE Difference >2

include_graphics(fig5CHIpath)

```

### 5. Summary Stats on CHI

```{r Summary Stats, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE}
source(here::here("scripts/statsCHI.r"), local = knitr::knit_global())

#tables-----
  #format threat contribution table
threats_tbl1 = threats_tbl[, c(1:4, 6:8)]
names(threats_tbl1) <- gsub('[.].+', '', names(threats_tbl1))
threats_tbl1[,1] <- gsub('[.]', ' ', threats_tbl1[,1])
threats_tbl1[,1] <- gsub('2', '', threats_tbl1[,1])

  kable(threats_tbl1, align=rep('l', 6), booktabs = TRUE, caption = "Summary of Impact by Threat",  digits = rep(2, 6))%>% add_header_above(c(" " = 1, "Early Period \n1988-2004" = 3, "Contemporary Period \n2005-2019" = 3)) %>%kable_styling(latex_options = "HOLD_position")%>% row_spec (0, bold = T) 
  
  #format CHI table

  summarytc1 = summarytc[, c(1, 4, 9,8, 6:7)]
  names(summarytc1) <- c("Period", "Median", "Mean","sd", "Max", "Sum")
summarytc1[,1] = c("Early 1988- 2004", "Contemporary 2005-2019")
  summarytc1%>%kable(align=rep('l', 7), booktabs = TRUE, caption = "Cumulative Impact Scores by Period", digits = rep(3, 6)) %>%kable_styling(latex_options = "HOLD_position")%>% row_spec (0, bold = T)
  
  round(RMSError, digits = 3)
  
#format dif table
  BigDif%>%kable(align=rep('c', 3), booktabs = TRUE, caption = "Diference in CHI increase and decrease 1988-2019", digits = rep(3, 6)) %>%kable_styling(latex_options = "HOLD_position")%>% row_spec (0, bold = T)

```
