---
title: "Cumulative Human Impact (CHI) scripts Methods"
author: "Feyrer et al. 2024"
date: "July 2024"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup , include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      fig_caption= TRUE,
                      fig.align = "left",
                      fig.pos = 'h',
                      dev = "png", 
                      cache = F
                )
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r libraries, include=FALSE}
#libraries
pacman::p_load(knitr, dplyr, sf, viridis, ggplot2, 
stars,
stringr,
here,
fasterize,
ggtext,
ggspatial,
pals,
tidyr,
fuzzyjoin,
kableExtra)

#read in functions
source(here::here("scripts/functionsCHI.r"))

options(scipen = 999) 

```

------------------------------------------------------------------------

## Data Processing {.tabset}

### 1. Import basemap data:

-   This includes Coastlines, bathymetry, NBW habitat area, and
    conservation zones
-   Code imports data and transforms to UTM Zone 20 where necessary
-   Extent of study area is based on area of influence around NBW
    habitat being a 50km buffer of the 1000m bathy line
-   Bathymetric data from Atlas of Canada available at:
    (<https://ftp.maps.canada.ca/pub/nrcan_rncan/vector/framework_cadre/North_America_Atlas10M/bathymetry/>)
-   Conservation zones compiled from layers available as shapefiles.
    -   Other Effective Conservation Measures (OECM) from
        <https://open.canada.ca/data/en/dataset/44769543-7a23-4991-a53f-c2cf7c7a946f>\
    -   Ocean's Act MPAs from
        <https://open.canada.ca/data/en/dataset/a1e18963-25dd-4219-a33f-1a38c4971250>\
    -   Georges Bank Oil and Gas Exclusion zone from CNSOPB
        <https://callforbids.cnsopb.ns.ca/2016/01/data-environment/gis-information>\
    -   Northern bottlenose whale Critical Habitat designations from
        Species at Risk
        <https://open.canada.ca/data/en/dataset/db177a8c-5d7d-49eb-8290-31e6a45d786c>

```{r import_basemap, include=TRUE, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 5, fig.fullwidth = TRUE}
source(here::here("scripts/baseMap.r"), local = knitr::knit_global())
source(here::here("scripts/functionsCHI.r"), local = knitr::knit_global())
#remove any previous results before knitting and rerunning all scripts
unlink(here::here("output/GRIDS/scaled/*"))
#map base map

include_graphics(gg_Fig2path)

```

------------------------------------------------------------------------

### 2. Map threats based on relative intensity of human activities

-   The spatial distribution of serious threats across the study area
    was interpreted from available information on the intensity of
    associated human activities: Entanglement, Depredation (Fishing
    effort), Military Sonar (Military Exercise Zones), Plastics (Density
    of large debris and micro plastics), Large Oil spills (Oil and Gas
    Development) and Vessel strikes (Vessel Traffic).
-   Whaling has not occurred in the region across the study period.
-   Data is divided into two periods, historical (1988-2004) and
    contemporary (2005-2023). Information on the intensity of human
    activities was standardized to values between 0-1.

#### A. Fishing Effort (Entanglement, Depredation)

-   Fishing effort was assessed using a combination of publicly
    available intensity data (time spent in each cell) for the region
    for three primary gear classes (i) Groundfish Fixed gear, e.g.,
    bottom longline, (ii) Pelagic longline, and (ii) Groundfish Mobile
    gear, e.g. trawl fisheries.
-   Differences in the methods for assessing effort between fisheries
    and time periods result in some differences in the resolution of
    available information. All effort data was standardized into
    percentiles.

```{r Groundfish - Fixed Gear (bottom longline), include = TRUE, fig.fullwidth = TRUE, message=FALSE, warning=FALSE}
source(here::here("scripts/fishMap.r"), local = knitr::knit_global())

#Bottom Longline
# Plot map

m2+m3+ plot_layout(guides = "collect")& theme(legend.position = 'bottom')

```

```{r Pelagic Longline, include = TRUE, fig.fullwidth = TRUE}

#Pelagic Longline
# Plot map

m4+m5+ plot_layout(guides = "collect")& theme(legend.position = 'bottom')

```

```{r Grounfish Mobile, include = TRUE, fig.fullwidth = TRUE}

#Early Period (1988-2004)------# # Contemporary Period (2005-2019)-------
# Plot map
m6 +m7 + plot_layout(guides = "collect")& theme(legend.position = 'bottom')

```

#### B. Military exercise zones (Military Sonar)

-   Military exercises– The Department of National Defence has
    designated Firing Practice and Exercise Areas off the coasts of
    Canada. Activities in these areas may include bombing practice from
    aircraft, air-to-air, air-to-sea or ground firing, and anti-aircraft
    firing, etc. In Atlantic Canada, this includes sea area employments
    for sub-surface operations and firing exercises (FIREX) (Department
    of National Defense Government of Canada, 2021). Polygon shapefiles
    of the designated areas were downloaded from the Canadian Government
    open data portal
    (<https://open.canada.ca/data/en/dataset/73111c78-298b-4be9-97f1-7aaa73cab477>).

-   With no effort data on military exercises in these areas across the
    entire period, we interpreted them to represent threat potential for
    military mid-frequency active sonar (MMFAS). Although noise impacts
    can extend outside these areas, the most acute threat would occur
    inside the exercise areas. According to the Canadian Coast Guard
    Navigational Warnings (NAVWARNs), previously known as Notices to
    Shipping (NOTSHIPS), there were 44 notices of military exercises in
    areas overlapping with beaked whale habitat, on 54 days in 2019 (the
    first year notices were made available online), lasting between 1-5
    days each (Government of Canada, 2021). However, this included 8
    (non-consecutive) days of the biannual international CUTLASS FURY
    exercises. This biannual event began in 2016 and has been described
    as the “largest international military exercise in Canadian waters”
    (The Canadian Press, 2016). Excluding CUTLASS FURY 2019, we used the
    proportion of days per year with notifications to mariners (46 days
    = 12.6%) as a baseline for effort across the area in any given year.
    Although there is little information on the extent of military
    activity in the early period, there were two large international
    exercises as part of the public record in 2016 and 2019 (i.e.,
    CUTLASS FURY). To account for the impact of the international
    exercises, we have increased the % effort to 15% per year in the
    period after 2004. In addition, we removed the MPA from this layer
    in the contemporary period, as the MPA management plan (Fisheries
    and Oceans Canada, 2017) indicates that although the Gully MPA
    cannot legally exclude military activities, in practise the
    government agrees that such exercises do not occur inside the MPA.

```{r DND, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE, fig.fullwidth = TRUE}

source(here::here("scripts/militaryexMap.r"), local = knitr::knit_global())

#Early Period (1988-2004)------# # Contemporary Period (2005-2019)-------
# Plot map
m8+m9 + plot_layout(guides = "collect")& theme(legend.position = 'bottom')

```

#### C. Oil and Gas Development (Large Oil spills)

-   Publicly available spatial and temporal data was collated from the
    Canada-Nova Scotia Offshore Petroleum Board (C-NSOPB) websites
    (CNSOPB, 2021). Shapefiles for existing pipelines, wells, platforms,
    exploratory licenses, significant discovery licenses, production
    licenses and sector announcements provided as part of the call for
    bids processes were downloaded, cleaned, and checked for
    duplication.

-   Activities occurring before and after 2004 were identified based on
    metadata in the shapefile or provided elsewhere on the website.

    -   Oil & Gas Operations includes wells, pipelines, and platforms
    -   Some infrastructure has been decommissioned since it was built,
        however the ongoing risk associated with pollution is assumed to
        continue over the entire study period.

```{r O&G operations, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE}

 source(here::here("scripts/oilngasMap.r"), local = knitr::knit_global())

#Early Period (1988-2004)------# # Contemporary Period (2005-2019)-------
# Plot maps
m10+ m11 + plot_layout(guides = "collect")& theme(legend.position = 'bottom')
```

#### D. Commercial Vessel Traffic (Vessel Strike)

-   Intensity raster of global shipping traffic for early period based
    on Scotian Shelf Human Use Atlas
-   Does not include all shipping data but provides a description of
    shipping patterns and intensity during 2002. Available data includes
    only inbound traffic; however, the density pattern of departing
    vessels is much the same as the inbound traffic pattern.
-   There is a voluntary vessels avoidance area around the Gully MPA,
    however there is no regulatory exclusion of this threat.

```{r Shipping Effort, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE}
source(here::here("scripts/shipMap.r"), local = knitr::knit_global())

m14

```

### 3. Assessing cumulative impacts from individual threats

-   Scale stressor intensity data so values range between 0-1 for layers
    with effort information: Oil & Gas Explorations, Fishing effort,
    Shipping, SST anomalies
-   Read in Pre/Post layers, Log10[x+1] transform and then normalize
    effort grids by max value across both periods
-   Presence /absence of stressor footprint data values should be either
    0 or 1
-   Stressor layers (effort or footprint) for each of the two periods
    are multiplied by the NBW vulnerability score
-   NBW Vulnerability score assigned based on sensitivity, temporal
    frequency and extent of stressor (See Impact Score Rationale below,
    Table 2 in Feyrer et al. 2022)
-   Finally the values of stressors in each period are summed to assess
    Cumulative impact

```{r Calc CHI, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE }
source(here::here("scripts/rescaleStressor.r"), local = knitr::knit_global())
source(here::here("scripts/calcCHI.r"), local = knitr::knit_global(), )

# plot by threat
include_graphics(fig3threatspath)
ImpactScore1 = ImpactScore[c(1,3,2, 4:6), 2:6]
names(ImpactScore1) <- gsub('[.].+', '', names(ImpactScore1))

#Table 1
 #format table
  kbl(ImpactScore1, booktabs = TRUE, caption = "Impact Score Rationale Table" ) %>%kable_styling(latex_options="scale_down")%>% row_spec (0, bold = T)%>%landscape( margin = NULL)

  
  #plot CHI PRE/ POST
include_graphics(fig4CHIpath)
```

### 4. Calculating difference in CHI between periods

-   The difference in Cumulative impact between periods highlights
    measures the cell-by-cell difference between CHI
-   Can be used to understand areas and intensity of differences in
    threats to NBW over time
-   Absolute diference is Contemporary period - Early Period

##### RMS Error

-   Square Root(sum(e\^2)/n)
-   About 2/3 of all the cells will differ by less than the rmse.
-   About 95% of all the cells will differ by less than twice the rmse.

```{r Calc CHI difference, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE}

# Plot Absolute Difference between periods
# Plot RMSE Difference & Plot RMSE Difference >2

include_graphics(fig5CHIpath)

```

### 5. Summary Stats on CHI

```{r Summary Stats, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE}
source(here::here("scripts/statsCHI.r"), local = knitr::knit_global())

#tables-----
  #format threat contribution table
threats_tbl1 = threats_tbl[, c(1:4, 6:8)]
names(threats_tbl1) <- gsub('[.].+', '', names(threats_tbl1))
threats_tbl1[,1] <- gsub('[.]', ' ', threats_tbl1[,1])
threats_tbl1[,1] <- gsub('2', '', threats_tbl1[,1])

  kable(threats_tbl1, align=rep('l', 6), booktabs = TRUE, caption = "Summary of Impact by Threat",  digits = rep(2, 6))%>% add_header_above(c(" " = 1, "Early Period \n1988-2004" = 3, "Contemporary Period \n2005-2019" = 3)) %>%kable_styling(latex_options = "HOLD_position")%>% row_spec (0, bold = T) 
  
  #format CHI table

  summarytc1 = summarytc[, c(1, 4, 9,8, 6:7)]
  names(summarytc1) <- c("Period", "Median", "Mean","sd", "Max", "Sum")
summarytc1[,1] = c("Early 1988- 2004", "Contemporary 2005-2019")
  summarytc1%>%kable(align=rep('l', 7), booktabs = TRUE, caption = "Cumulative Impact Scores by Period", digits = rep(3, 6)) %>%kable_styling(latex_options = "HOLD_position")%>% row_spec (0, bold = T)
  
  round(RMSError, digits = 3)
  
#format dif table
  BigDif%>%kable(align=rep('c', 3), booktabs = TRUE, caption = "Diference in CHI increase and decrease 1988-2019", digits = rep(3, 6)) %>%kable_styling(latex_options = "HOLD_position")%>% row_spec (0, bold = T)

```
